Predicting Conflict Escalation

### 2023-05-03 docs/hes_final/article.qmd
add references, fix figure captions, rm line breaks

---
title: "Predicting Conflict Escalation"
author:
  - name: Diana Kolusheva
    email: email@h
    affiliation: Harvard University
  - name: Chad Stoughton
    email: email@h
    affiliation: Harvard University
  - name: Evan Wheeler
    email: email@h
    affiliation: Harvard University
lang: en
format:
  pdf:
    classoption:
    - twocolumn
    margin-left: 25mm
    margin-right: 25mm
    margin-top: 25mm
    margin-bottom: 25mm
    whitespace: medium
bibliography: bibliography.bibtex
# https://github.com/quarto-dev/quarto-cli/discussions/2786
include-in-header:
  - text: |
      \usepackage{float}
      \makeatletter
      \let\oldlt\longtable
      \let\endoldlt\endlongtable
      \def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
      \def\longtable@i[#1]{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt[#1]
      }
      \def\longtable@ii{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt
      }
      \def\endlongtable{\endoldlt
      \end{minipage}
      \twocolumn
      \end{figure}}
      \makeatother
# https://tex.stackexchange.com/questions/31301/how-to-reduce-the-number-of-hyphenation
include-before-body:
  - text: |
      \pretolerance=5000
      \tolerance=9000
      \emergencystretch=0pt
      \exhyphenpenalty=100
      \righthyphenmin=4
      \lefthyphenmin=5
      \setlength{\belowcaptionskip}{-5pt}
embed-resources: true
notebook-links: false
title-block-banner: true
abstract: |
  The task of forecasting conflict escalation is of great interest to policy makers, executives, and social scientists. In this article, we propose a novel, text-first approach to computational conflict prediction using contemporary natural language processing and machine learning techniques. We outline three methods to extract generalizable conflict signals directly from a corpus of ofer 2.5 million document embeddings derived from publicly available news sources. The first uses text data enriched by country context features, the second uses text data conditioned only on the country label itself, and the third extracts a signal from the text alone, with no additional country or region-specific features. We demonstrate that usable conflict signals can be extracted directly from text sources, and we outline potential avenues for further research that we believe may yield models that can provide early warnings to decision makers to facilitate strategic planning, resource allocation, and disaster mitigation.

@@

We developed two approaches for extracting conflict signals from text embeddings. In the first approach we used a fit-in-memory dataset of monthly embedding averages per country per month and applied statistical classification models such as K Nearest Neighbors and XGBoost. In the second approach we stacked multiple article-level embeddings and applied Transformer-based models to extract a signal directly from the text corpus. See the figures (X, Y) for details of both methods.

@@

![The Spike metric applied to Ukraine](imgs/Ukraine_Spikes.png)

Figure 1 shows how this metric captures conflict fluctuations in Ukraine. Prior to the Russian invasion of February 2022, small fluctuations were able to trigger a spike. However, much larger fluctuations in the period following the invasion are filtered out, as the spike formula adjusts to the higher level of baseline conflict. The increase in violence during the Ukrainian counter-offensive in the late summer again triggers a spike, and this reflects significant changes in the situation on the ground.

@@

```{mermaid}
%%| fig-width: 3
%%| fig-height: 5
%%| fig-cap: Fig 1. Overview of data generation process for our conflict dataset 
flowchart TD
    GSG[GDELT Global Similarty Graph] -->|sample 20% of daily files| Merge{Merge on article URL,\n select relevant features}
    GKG[GDELT Global Knowledge Graph] -->|sample 20% of daily files| Merge

@@

    GDELT --> Data(Time-partitioned train, val, & test datasets)
    Spike --> Data
```
### Conflict Prediction Using Text and Country Features

@@

![Distribution of supervised UMAP reduction to 2 dimensions](imgs/fig-umap2d-jointplot.png){width=50%}

![Scatterplot of 2d UMAP reduced articles with spike labels](imgs/fig-umap2d-ds-targets.png){width=50%}

Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total') and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges, 4 categories of gross national income per capita, and 3 categories of Fragile States Index ranges).

```{mermaid}
%%| fig-width: 3
%%| fig-height: 3
%%| fig-cap: Fig 6. Feed forward neural network model architecture
flowchart TD
    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)

@@

    Context[Country context indicators] --> FFNN
```
### Conflict Prediction Using Text and Country Labels

@@

```{mermaid}
%%| fig-width: 3
%%| fig-height: 3
%%| fig-cap: Fig 2. Statistical model architecture
flowchart TD
    GDELT[GDELT Embeddings] -->|average by country by month| AGG(Monthly news representation)
    Countries[Country labels] -->|one-hot encoding| AGG

@@

    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
    Spike(Conflict response\n binary variable) --> XGD
    Spike --> KNN
```
### Conflict Prediction Using Text Alone

@@

![Reducing Embedding Dimensionality with Transformer Encoder](imgs/Transformer_Viz.png){width=50%}

@@

```{mermaid}
%%| fig-width: 3
%%| fig-height: 3
%%| fig-cap: Fig 3. Transformer model architecture
flowchart TD
    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)

@@

    Spike -->|included during model training| Trans1
```
## Results

@@

![Model performance metrics (higher values are better)](imgs/tbl-best-max-metrics.png){width=50%}

![Model performance metrics (lower values are better)](imgs/tbl-best-min-metrics.png){width=50%}

![Receiver Operator Characteristics](imgs/fig-roc.png){width=50%}

![Precision Recall Curve](imgs/fig-prc.png){width=50%}

@@

![Transformer model Predictions on test data](imgs/fig-transformer-test-predictions.png){width=55%}

![XGBoost model predictions on test data](imgs/fig-xgboost-test-predictions.png){width=55%}

![FFNN model predictions on test data](imgs/fig-ffnn-test-predictions.png){width=55%}

>>>>

---
title: "Predicting Conflict Escalation"
author:
  - name: Diana Kolusheva
    email: email@h
    affiliation: Harvard University
  - name: Chad Stoughton
    email: email@h
    affiliation: Harvard University
  - name: Evan Wheeler
    email: email@h
    affiliation: Harvard University
lang: en
format:
  pdf:
    classoption:
    - twocolumn
    margin-left: 25mm
    margin-right: 25mm
    margin-top: 25mm
    margin-bottom: 25mm
    whitespace: medium
    mermaid:
      theme: neutral
bibliography: bibliography.bibtex
# https://github.com/quarto-dev/quarto-cli/discussions/2786
include-in-header:
  - text: |
      \usepackage{float}
      \makeatletter
      \let\oldlt\longtable
      \let\endoldlt\endlongtable
      \def\longtable{\@ifnextchar[\longtable@i \longtable@ii}
      \def\longtable@i[#1]{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt[#1]
      }
      \def\longtable@ii{\begin{figure}[H]
      \onecolumn
      \begin{minipage}{0.5\textwidth}
      \oldlt
      }
      \def\endlongtable{\endoldlt
      \end{minipage}
      \twocolumn
      \end{figure}}
      \makeatother
# https://tex.stackexchange.com/questions/31301/how-to-reduce-the-number-of-hyphenation
include-before-body:
  - text: |
      \pretolerance=5000
      \tolerance=9000
      \emergencystretch=0pt
      \exhyphenpenalty=100
      \righthyphenmin=4
      \lefthyphenmin=5
      \setlength{\belowcaptionskip}{-5pt}
embed-resources: true
notebook-links: false
title-block-banner: true
abstract: |
  The task of forecasting conflict escalation is of great interest to policy makers, executives, and social scientists. In this article, we propose a novel, text-first approach to computational conflict prediction using contemporary natural language processing and machine learning techniques. We outline three methods to extract generalizable conflict signals directly from a corpus of ofer 2.5 million document embeddings derived from publicly available news sources. The first uses text data enriched by country context features, the second uses text data conditioned only on the country label itself, and the third extracts a signal from the text alone, with no additional country or region-specific features. We demonstrate that usable conflict signals can be extracted directly from text sources, and we outline potential avenues for further research that we believe may yield models that can provide early warnings to decision makers to facilitate strategic planning, resource allocation, and disaster mitigation.

@@

We developed three approaches for extracting conflict signals from text embeddings. The first approach used supervised dimensionality reduction and a feed-forward neural network conditioned with country context characteristics. In the second approach we used a fit-in-memory dataset of monthly embedding averages per country per month and applied statistical classification models such as K Nearest Neighbors and XGBoost. In the third approach we stacked multiple article-level embeddings and applied Transformer-based models to extract a signal directly from the text corpus. See the @fig-ffnn, @fig-xgboost, and @fig-transformer for details of modeling approaches.

@@

![The Spike metric applied to Ukraine](imgs/Ukraine_Spikes.png){#fig-spike}

@fig-spike shows how this metric captures conflict fluctuations in Ukraine. Prior to the Russian invasion of February 2022, small fluctuations were able to trigger a spike. However, much larger fluctuations in the period following the invasion are filtered out, as the spike formula adjusts to the higher level of baseline conflict. The increase in violence during the Ukrainian counter-offensive in the late summer again triggers a spike, and this reflects significant changes in the situation on the ground.

@@

::: {#fig-dgp layout-ncol=1}
```{mermaid}
%%| fig-width: 3
%%| fig-height: 5
flowchart TD
    GSG[GDELT Global Similarty Graph] -->|sample 20% of daily files| Merge{Merge on article URL,\n select relevant features}
    GKG[GDELT Global Knowledge Graph] -->|sample 20% of daily files| Merge

@@

    GDELT --> Data(Time-partitioned train, val, & test datasets)
    Spike --> Data
```
Overview of data generation process for our conflict dataset 
:::

### Conflict Prediction Using Text and Country Features

@@

![Density of embeddings resulting from supervised UMAP reduction to 2 dimensions](imgs/fig-umap2d-jointplot.png){#fig-umap2d-jointplot}

![Scatterplot of 2-dimension UMAP-reduced articles with spike labels](imgs/fig-umap2d-ds-targets.png){#fig-umap2d-ds-targets}

Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total' from UNICEF's State of the World's Children report) and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges [@hdi], 4 categories of gross national income per capita [@wb], and 3 categories of Fragile States Index ranges[@fsi]).

::: {#fig-ffnn layout-ncol=1}
```{mermaid}
%%| fig-width: 3
%%| fig-height: 3
flowchart TD
    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)

@@

    Context[Country context indicators] --> FFNN
```
Feed forward neural network model architecture
:::
### Conflict Prediction Using Text and Country Labels

@@

::: {#fig-xgboost layout-ncol=1}
```{mermaid}
%%| fig-width: 3
%%| fig-height: 3
flowchart TD
    GDELT[GDELT Embeddings] -->|average by country by month| AGG(Monthly news representation)
    Countries[Country labels] -->|one-hot encoding| AGG

@@

    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
    Spike(Conflict response\n binary variable) --> XGD
    Spike --> KNN
```
Statistical model architecture
:::
### Conflict Prediction Using Text Alone

@@

![Reducing Embedding Dimensionality with Transformer Encoder](imgs/Transformer_Viz.png){#fig-transformer-viz}

@@

::: {#fig-transformer layout-ncol=1}
```{mermaid}
%%| fig-width: 3
%%| fig-height: 3
flowchart TD
    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)

@@

    Spike -->|included during model training| Trans1
```
Transformer model architecture
:::
## Results

@@

![Model performance metrics (higher values are better)](imgs/tbl-best-max-metrics.png){#fig-best-max-metrics}

![Model performance metrics (lower values are better)](imgs/tbl-best-min-metrics.png){#fig-best-min-metrics}

![Receiver Operator Characteristics](imgs/fig-roc.png){#fig-roc}

![Precision Recall Curve](imgs/fig-prc.png){#fig-prc}

@@

![FFNN model predictions on test data](imgs/fig-ffnn-test-predictions.png){#fig-ffnn-test-predictions}

![XGBoost model predictions on test data](imgs/fig-xgboost-test-predictions.png){#fig-xgboost-test-predictions}

![Transformer model Predictions on test data](imgs/fig-transformer-test-predictions.png){#fig-transformer-test-predictions}
